---
title: "exercise-12"
---
Keerthana Stanley



Preliminaries: 

loading in the data
```{r, message=FALSE}
library(tidyverse)

f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt"

d <- read_tsv(f, col_names=TRUE)
```


exploratory data analysis using {skimr}
```{r}
library(skimr)

d_sumstats <- skim(d)
```


QUESTION:
Which of the variables are categorical and which are numeric?

ANSWER:
  - categorical: order, Family, Genus, species
  - numerical: mass, gestation, newborn, weaning, wean mass, AFR, max.life, litter size, litters/year, refs
  
  









STEP 1:
Replace all values of -999 (the authors’ code for missing data) with NA.

documentation for using the {naniar} for this : https://www.rdocumentation.org/packages/naniar/versions/1.1.0/topics/replace_with_na_all


```{r, message=FALSE}
library(naniar)

# replace_with_na_all(dataset, condition)
d <- replace_with_na_all(d, condition = ~.x == -999)
```






STEP 2
Drop the variables litter size and refs.
```{r}
d <- d %>% 
  # using select() to pull out specific columns
  # but in this case I'm using the '-' to exclude litter size and refs
  select(-c(`litter size`, `refs`))
```







STEP 3:
Log transform all of the other numeric variables.


mutate(across(where(), .funs)) allows me to apply one function to multiple columns:
https://dplyr.tidyverse.org/reference/across.html

my code is based more off of code here: https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-colwise/



so I wanted to mutate WHERE the variable is numeric and add a log function
```{r}
# using pipe operator for this
d_log <- d %>%
  # where-- is.numeric; function-- log
  mutate(across(where(is.numeric), log))
```

now gestation(mo), weaning(mo), AFR(mo), max. life(mo), newborn(g), wean mass, and (overall body) mass are all actually the log versions of their original value






STEP 4

Regress the (now log transformed) age [gestation(mo), weaning(mo), AFR(mo) (age at first reproduction), and max. life(mo) (maximum lifespan)] and mass [newborn(g) and wean mass(g)] variables on (now log transformed) overall body mass(g) 

Add the residuals to the dataframe as new variables (relGest, relWean, relAFR, relLife, relNewbornMass, and relWeaningMass).


**** Use “na.action=na.exclude” in your lm() calls!!! 
With this argument set, the residuals will be padded to the correct length by inserting NAs for cases with missing data.


once again using mutate() for these changes/additions in the dataset


each variable listed above is regressed to  mass(g)
first just printing out the summary for each regression model:
```{r}
m_gestation <- lm(`gestation(mo)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_gestation)

m_weaning <- lm(`weaning(mo)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_weaning)

m_AFR <- lm(`AFR(mo)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_AFR)

m_max_life <- lm(`max. life(mo)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_max_life)

m_newborn <- lm(`newborn(g)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_newborn)

m_wean_mass <- lm(`wean mass(g)` ~ `mass(g)`, data = d_log, na.action = na.exclude)
summary(m_wean_mass)
```



```{r}
d_new <- d_log %>%
  mutate(
    # gestation 
    relGest = resid(lm(`gestation(mo)` ~ `mass(g)`, data = ., na.action = na.exclude)),
    # weaning 
    relWean = resid(lm(`weaning(mo)` ~ `mass(g)`, data = ., na.action = na.exclude)),
    # AFR
    relAFR = resid(lm(`AFR(mo)` ~ `mass(g)`, data = ., na.action = na.exclude)),
    # max life 
    relLife = resid(lm(`max. life(mo)` ~ `mass(g)`, data = ., na.action = na.exclude)),
    # newborn
    relNewbornMass = resid(lm(`newborn(g)` ~ `mass(g)`, data = ., na.action = na.exclude)),
    # wean mass
    relWeaningMass = resid(lm(`wean mass(g)` ~ `mass(g)`, data = ., na.action = na.exclude))
  )

```









STEP 5:
- Plot residuals of max lifespan (relLife) in relation to Order. 

Q: Which mammalian orders have the highest residual lifespan?
A: primates


- Plot residuals of newborn mass (relNewbornMass) in relation to Order. 

Q: Which mammalian orders have the have highest residual newborn mass?
A: macroscelidia

- Plot residuals of weaning mass (relWeaningMass) in relation to Order. 

Q: Which mammalian orders have the have highest relative weaning mass?
A: perissodactyla 



using ggplot2 for the plotting
```{r}
library(ggplot2)

(p1_relLife <- ggplot(d_new, aes(x = order, y = relLife)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Residuals of Max Lifespan by Order", 
       x = "Order", 
       y = "Residual Max Lifespan"))

(p2_relNewbornMass <- ggplot(d_new, aes(x = order, y = relNewbornMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Residuals of Newborn Mass by Order", 
       x = "Order", 
       y = "Residual Newborn Mass"))

(p3_relWeaningMass <- ggplot(d_new, aes(x = order, y = relWeaningMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Residuals of Weaning Mass by Order", 
       x = "Order", 
       y = "Residual Weaning Mass"))

```




